---
title: "Variation in E. coli Measurements at Toronto Beaches"
author: 
  - Siyi Zhu
thanks: "Code and data are available at: https://github.com/Sophiaaa12/toronto-beaches-water-quality."
date: today
date-format: long
abstract: "This paper examines variation in E. coli measurements at Toronto beaches using 102,534 observations collected from 2007 to 2026. The measurements are strongly right-skewed, with most observations concentrated at relatively low values and a small number of extremely high values. E. coli measurements also differ across beaches, with Sunnyside Beach and Marie Curtis Park East Beach generally having higher values than several other beaches. Measurements vary over time, but there is no clear sustained upward or downward pattern over the study period."
format:
  pdf:
    include-in-header:
      text: |
        \usepackage{float}
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(tinytable)
analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))
```

# Introduction

Toronto’s public beaches provide important recreational spaces during the summer swimming season, but water quality can vary across locations and over time. One commonly used indicator of recreational water quality is the concentration of *Escherichia coli* (E. coli), which is monitored by Toronto Public Health to assess conditions at supervised beaches [@torontobeaches]. E. coli is also used as an indicator of fecal contamination in recreational fresh water [@healthcanada]. Understanding how E. coli measurements are distributed and how they differ across beaches and over time can help describe variation in recreational water quality across Toronto.

Using Toronto Beaches Water Quality data from 2007 to 2026, this paper examines the distribution of E. coli measurements across Toronto beaches. The distribution of E. coli levels is strongly right-skewed: most observations are concentrated at relatively low values, while a small proportion of observations have extremely high values. E. coli levels also differ across beaches, with Sunnyside Beach and Marie Curtis Park East Beach generally having higher values than several other beaches. Recent Canadian research has also examined the relationship between E. coli levels and acute gastrointestinal illness among beachgoers [@Young2026].

The remainder of this paper is structured as follows. @sec-data describes the Toronto Beaches Water Quality dataset, explains how the E. coli measurements are collected and recorded, and examines their overall distribution and variation across beaches and over time.

# Data {#sec-data}

## Overview

The Toronto Beaches Water Quality dataset is provided through the City of Toronto Open Data Portal [@torontobeaches]. The data were downloaded reproducibly using the `opendatatoronto` R package [@opendatatoronto]. During the summer swimming season, the City of Toronto collects water samples from supervised public beaches, and Toronto Public Health analyzes the samples for E. coli as part of its monitoring of beach water quality [@torontobeaches].

The data used in this analysis contain 102,534 observations collected between June 3, 2007 and September 16, 2026 across 10 Toronto beaches. Each observation records a beach identifier, beach name, sampling site, collection date, and E. coli measurement. The original dataset also contains an observation identifier and geographic information, which are not required for this analysis and were removed during data cleaning. Observations with missing E. coli measurements were retained in the cleaned dataset to preserve the original observations but excluded from subsequent analyses to ensure that only observed E. coli values were analyzed. The analysis was conducted using R [@R], the `tidyverse` [@tidyverse], and the `here` package [@here]. Figures were produced using `ggplot2` [@ggplot], and tables were produced using `tinytable` [@tinytable].

## Measurement

Each row in the analysis data represents an E. coli measurement from a particular sampling site at a Toronto beach on a particular collection date. Multiple observations can be recorded for the same beach and date because samples are collected at different sites. Therefore, an observation should not be interpreted as the overall water quality of an entire beach on that day.

@tbl-data-summary summarizes the distribution and the number of missings of E. coli measurements in the analysis data, showing that the measurements are strongly right-skewed, with a median of 20 and a mean of approximately 154. The largest recorded value is 6,191,768, illustrating the presence of extreme observations. Because the mean is sensitive to these unusually large values, the median provides a more representative description of a typical observation. In addition, 3,543 E. coli measurements are missing. As described above, these observations are retained in the cleaned dataset and excluded only from calculations or visualizations that require an observed E. coli value.

```{r}
#| label: tbl-data-summary
#| tbl-cap: "Summary of E. coli measurements in the analysis data"
#| echo: false
#| warning: false
#| message: false

data_summary <- tibble(
  Statistic = c(
    "Number of observations",
    "Number of beaches",
    "Median E. coli",
    "Mean E. coli",
    "Minimum E. coli",
    "Maximum E. coli",
    "Missing E. coli"
  ),
  Value = c(
    format(nrow(analysis_data), big.mark = ","),
    n_distinct(analysis_data$beachName),
    median(analysis_data$eColi, na.rm = TRUE),
    round(mean(analysis_data$eColi, na.rm = TRUE)),
    min(analysis_data$eColi, na.rm = TRUE),
    format(max(analysis_data$eColi, na.rm = TRUE), big.mark = ","),
    format(sum(is.na(analysis_data$eColi)), big.mark = ",")
    )
  )

tt(data_summary)
```

There are limited direct privacy concerns because the dataset contains environmental measurements rather than information about individuals. However, the measurements have public-health relevance and should therefore be interpreted carefully. Individual site-level observations are not classified as “safe” or “unsafe” in this analysis because they do not necessarily represent the water quality of an entire beach.

## Data exploration

### Overall distribution

@fig-ecoli-distribution shows the overall distribution of E. coli measurements. A logarithmic scale is used on the horizontal axis to display the wide range of observed values more clearly. The figure displays measurements up to 10,000 so that the distribution of the large majority of observations remains visible. Measurements above this value are retained in the analysis data and included in the numerical summaries in @tbl-data-summary.

```{r}
#| label: fig-ecoli-distribution
#| fig-cap: "Overall distribution of E. coli measurements up to 10,000"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 5
#| out-width: "85%"
#| fig-align: center
#| fig-pos: H

analysis_data |>
  filter(
    !is.na(eColi),
    eColi <= 10000
  ) |>
  ggplot(aes(x = eColi)) +
  geom_histogram(
    bins = 50
  ) +
  scale_x_log10() +
  labs(
    x = "E. coli measurement (log scale)",
    y = "Number of observations"
  ) +
  theme_minimal()
```

As shown in @fig-ecoli-distribution, the distribution is strongly right-skewed, with most observations concentrated at relatively low values and progressively fewer observations at higher values. The long upper tail is consistent with the large difference between the median and mean reported in @tbl-data-summary.


### Variation across beaches

E. coli measurements differ across Toronto beaches, as shown in @fig-beach-distribution. Each point represents an individual E. coli measurement, and a logarithmic scale is used to display the wide range of observed values more clearly.

```{r}
#| label: fig-beach-distribution
#| fig-cap: "E. coli measurements across Toronto beaches"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 5
#| out-width: "90%"
#| fig-align: center
#| fig-pos: H

analysis_data |>
  filter(!is.na(eColi)) |>
  ggplot(
    aes(
      x = reorder(beachName, eColi, FUN = median),
      y = eColi
    )
  ) +
  geom_jitter(
    alpha = 0.08,
    width = 0.2,
    size = 0.4
  ) +
  scale_y_log10() +
  coord_flip() +
  labs(
    x = "Beach",
    y = "E. coli measurement (log scale)"
  ) +
  theme_minimal()
```

@fig-beach-distribution shows that Sunnyside Beach and Marie Curtis Park East Beach have more observations extending into higher E. coli values, while several other beaches have measurements more concentrated at lower values. However, E. coli measurements overlap considerably across beaches, with relatively high values observed at all beaches. This suggests that there are some differences in E. coli measurements between beaches despite the considerable overlap. Within individual beaches, E. coli measurements also vary substantially, with observations spanning a wide range of values.


### Variation over time

E. coli measurements also vary over time, as shown in @fig-time-distribution. Each point represents an individual E. coli measurement collected between 2007 and 2026. A logarithmic scale is again used to display the wide range of observed values more clearly.

```{r}
#| label: fig-time-distribution
#| fig-cap: "E. coli measurements over time"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 5
#| out-width: "90%"
#| fig-align: center
#| fig-pos: H

analysis_data |>
  filter(!is.na(eColi)) |>
  ggplot(aes(
    x = collectionDate,
    y = eColi
  )) +
  geom_point(
    alpha = 0.08,
    size = 0.4
  ) +
  scale_y_log10() +
  labs(
    x = "Collection date",
    y = "E. coli measurement (log scale)"
  ) +
  theme_minimal()
```

@fig-time-distribution shows substantial variation in E. coli measurements throughout the study period. Most observations are concentrated at relatively low values, while higher measurements occur in each year. One extreme observation appears around 2009, reaching close to $10^7$. Overall, there is no clear sustained upward or downward pattern in E. coli measurements over time.


# References
